---
title: "Blockforest_binario"
author: "Lorena"
date: "2024-05-13"
output: html_document
---

## BLOCKFOREST ONLY WITH DIFFERENTIATED AND PROLIFERATIVE, THE BEST DIFFERENTIATED VARIABLES

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
```

```{r}
library(blockForest)
library(caret)
library(dplyr)

# Load the omics data
load("C:/Users/Usuario/Desktop/CIENCIA DE DATOS/TERCERO/SEGUNDO CUATRI/PROYIII/TCGAforMORE.RData")
load("C:/Users/Usuario/Desktop/CIENCIA DE DATOS/TERCERO/SEGUNDO CUATRI/PROYIII/folds.RData")
load("C:/Users/Usuario/Desktop/CIENCIA DE DATOS/TERCERO/SEGUNDO CUATRI/PROYIII/train_indices.RData")

nombres_filas_TF <- rownames(omicas$TF)
nuevos_nombres_filas <- paste0(nombres_filas_TF, 'TF')
rownames(omicas$TF) <- nuevos_nombres_filas

# Filter patients that are only of subtype 'differentiated' and 'proliferative'
subtypes_filtered <- subtypes %>%
  filter(subtype %in% c("differentiated", "proliferative"))

# Filter the omics data matrices to match the filtered patients
pacientes_filtrados <- rownames(subtypes_filtered)

omicas_cnv_filtered <- omicas$CNV[, pacientes_filtrados, drop = FALSE]
omicas_methyl_filtered <- omicas$Methyl[, pacientes_filtrados, drop = FALSE]
omicas_miRNA_filtered <- omicas$miRNA[, pacientes_filtrados, drop = FALSE]
omicas_tf_filtered <- omicas$TF[, pacientes_filtrados, drop = FALSE]

# Create a list of blocks
blocks <- list("A" = 1:ncol(omicas_cnv_filtered), 
               "B" = (ncol(omicas_cnv_filtered) + 1):(ncol(omicas_cnv_filtered) + ncol(omicas_methyl_filtered)),
               "C" = (ncol(omicas_cnv_filtered) + ncol(omicas_methyl_filtered) + 1):(ncol(omicas_cnv_filtered) + ncol(omicas_methyl_filtered) + ncol(omicas_miRNA_filtered)),
               "D" = (ncol(omicas_cnv_filtered) + ncol(omicas_methyl_filtered) + ncol(omicas_miRNA_filtered) + 1):ncol(omicas_tf_filtered))

# Combine all filtered omics data into a single matrix
omicas_all_filtered <- rbind(omicas_cnv_filtered, omicas_methyl_filtered, omicas_miRNA_filtered, omicas_tf_filtered)

# Define X and Y for data splitting
X <- t(omicas_all_filtered)
subtypes_filtered$subtype <- factor(subtypes_filtered$subtype)
Y <- subtypes_filtered$subtype

# Split the dataset into training and test sets
set.seed(123)  # For reproducibility
#indices <- createDataPartition(Y, p = 0.8, list = FALSE)
x_train <- X[train_indices, ]
x_test <- X[-train_indices, ]
y_train <- Y[train_indices]
y_test <- Y[-train_indices]
```

## Hyperparameters
- mtry: A vector of length equal to the number of blocks, where the m-th entry of the vector gives the number of variables to sample from block m. The default values are sqrt(p_1) + sqrt(p_2) + ... + sqrt(p_M) where p_m denotes the number of variables in the m-th block (m = 1, ..., M) and sqrt() denotes the square root function. Let the function calculate it by default.
- splitrule, block.method, replace: Default values.
-> Test values for two hyperparameters: nsets and num.trees:
      - num.trees = c(100, 200)
      - nsets = c(100, 200)

```{r}
# Define the evaluation function for cross-validation
cv_func <- function(num.trees, nsets, folds, x_train, y_train, blocks) {
  cv_results <- lapply(folds, function(fold_index) {
    x_train_fold <- x_train[fold_index[[1]], ]
    y_train_fold <- y_train
    
    # Fit the blockForest model with the current fold
    result <- blockfor(X = t(x_train_fold), y = y_train_fold, blocks = blocks, 
                       num.trees = num.trees, replace = TRUE,
                       nsets = nsets, num.trees.pre = 50, splitrule = "gini", 
                       block.method = "BlockForest", importance = 'impurity')
  
    return(result)
  })
  
  return(cv_results)
}

# Define the ranges of values for the hyperparameters
param_grid <- expand.grid(
  num.trees = c(100, 200),
  nsets = c(10, 20)
)

# Apply cv_func for each combination of hyperparameters in param_grid
results <- mapply(function(num_trees, nsets) {
  cv_func(num.trees = num_trees, 
          nsets = nsets, 
          folds = folds, 
          x_train = x_train, 
          y_train = y_train, 
          blocks = blocks)
}, param_grid$num.trees, param_grid$nsets)
```

```{r}
# Define a function to get the minimum prediction error from the results
get_min_prediction_error <- function(results) {
  min_errors <- sapply(results, function(result) {
    min(result$forest$prediction.error)
  })
  return(min(min_errors))
}

# Filter the results to remove those with non-finite prediction errors
filtered_results <- results[sapply(results, function(result) all(is.finite(result$forest$prediction.error)))]

# Get the minimum prediction error from all filtered results
min_error <- get_min_prediction_error(filtered_results)

# Find the combination of hyperparameters with the minimum prediction error
best_combination_index <- which.min(sapply(filtered_results, function(result) min(results$forest$prediction.error)))
best_combination <- param_grid[best_combination_index, ]

print(paste("The hyperparameter combination with the lowest prediction error is:", 
            paste(names(best_combination), best_combination, sep = "=", collapse = ", "), 
            "with a prediction error of", min_error))
```

## Apply Model with Best Hyperparameters

```{r}
# Apply the blockfor model with the best combination of parameters on the test set
best_model <- blockfor(X = x_train, y = y_train, blocks = blocks, 
                       num.trees = 100, replace = TRUE,
                       nsets = 10, num.trees.pre = 50, splitrule = "gini", 
                       block.method = "BlockForest", importance = 'impurity')

# Convert the x_test matrix to a dataframe
x_test_df <- data.frame(x_test)

predictions <- predict(object = best_model$forest,
                       data = x_test_df, 
                       predict.all = FALSE, 
                       num.trees = 100, 
                       type = 'response', 
                       num.threads = 1, 
                       verbose = TRUE)

# Calculate the accuracy by comparing the predictions with the actual labels
accuracy <- mean(predictions$predictions == y_test)

# Display the result
print(paste("The accuracy of the model on the test set is:", accuracy))
```
```{r}
# Genera la matriz de confusión
conf_matrix <- confusionMatrix(predictions$predictions, y_test)

# Imprime la matriz de confusión
print(conf_matrix)
```

```{r}
best_model$paramvalues
```

## Top 10 Most Important Variables

```{r}
# Get the importance of variables
variable_importance <- best_model$forest$variable.importance

# Sort the variables by importance from highest to lowest
sorted_importance <- variable_importance[order(variable_importance, decreasing = TRUE)]

# Filter variables with positive importance
positive_importance <- sorted_importance[sorted_importance > 0]

# Filter variables with negative importance
negative_importance <- sorted_importance[sorted_importance < 0]

# Select the top 10 most important variables
top_10_importance <- sorted_importance[1:10]

# Check the column names in x_train
colnames_x_train <- colnames(x_train)

# Get the names of the top 10 most important variables
top_10_names <- names(top_10_importance)

# Check if the variable names are in x_train
top_10_names <- top_10_names[top_10_names %in% colnames_x_train]

# Create a bar plot of the top 10 most important variables with names
barplot(top_10_importance[top_10_names], main = "Top 10 Important Variables",
        xlab = "Importance", ylab = "Variables", names.arg = top_10_names, las = 2)
```

## Bottom 10 Least Important Variables

```{r}
# Sort the variables by importance from lowest to highest
sorted_importance <- variable_importance[order(variable_importance)]

# Select the bottom 10 least important variables
bottom_10_importance <- sorted_importance[1:10]

bottom_10_names <- names(bottom_10_importance)

# Create a bar plot of the bottom 10 least important variables
barplot(bottom_10_importance, main = "Bottom 10 Important Variables",
        xlab = "Variables", ylab = "Importance")
```

## Find Block for Each Important Variable

```{r}
# Get the names of variables for each block
names_cnv <- colnames(t(omicas_cnv_filtered))
names_methyl <- colnames(t(omicas_methyl_filtered))
names_miRNA <- colnames(t(omicas_miRNA_filtered))
names_tf <- colnames(t(omicas_tf_filtered))

# Create a list with the names of the variables and their corresponding block
block_names <- list(
  CNV = names_cnv,
  Methylation = names_methyl,
  miRNA = names_miRNA,
  TF = names_tf
)

# Function to find the block to which a variable belongs
find_block <- function(variable, block_names) {
  for (block in names(block_names)) {
    if (variable %in% block_names[[block]]) {
      return(block)
    }
  }
  return(NA)  # Return NA if not found in any block
}

# Important variables
important_vars <- c("SLC45A1", "SLC25A33", "RERE", "ERRFI1", "SLC2A7", "PARK7", "LINC00337", "MIR34A", "ESPN", "DFFB", "KCNAB2", "RNA5SP40", "CA6", "SLC2A5", "TNFRSF9", "PER3", "ATAD3B", "TMEM201")

# Create a dataframe to store important variables and their corresponding blocks
important_vars_blocks <- data.frame(
  Variable = important_vars,
  Block = sapply(important_vars, find_block, block_names = block_names)
)

print(important_vars_blocks)
```

## ANOVA

```{r}
# Create a table of the most important variables and their values in the training dataset
top_variables_data <- as.data.frame(x_train[, top_10_names])
top_variables_data$subtype <- y_train

# Analysis of the association between variables and cancer subtypes using ANOVA
anova_results <- list()

for (variable in top_10_names) {
  formula <- as.formula(paste(variable, "~ subtype"))
  anova_results[[variable]] <- summary(aov(formula, data = top_variables_data))
}

# Display ANOVA results
anova_summary <- sapply(anova_results, function(x) x[[1]][["Pr(>F)"]][1])
anova_summary <- data.frame(Variable = names(anova_summary), PValue = anova_summary)
anova_summary <- anova_summary %>% arrange(PValue)

print("ANOVA results for the most important variables:")
print(anova_summary)
```

