---
title: "prueba"
author: "Lorena"
date: "2024-05-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
```

## BLOCKFOREST MODEL WITH ALL THE DATASET

```{r}
library(blockForest)
library(caret)
library(dplyr)

# Load omics data
load("C:/Users/Usuario/Desktop/CIENCIA DE DATOS/TERCERO/SEGUNDO CUATRI/PROYIII/TCGAforMORE.RData")
load("C:/Users/Usuario/Desktop/CIENCIA DE DATOS/TERCERO/SEGUNDO CUATRI/PROYIII/folds.RData")

nombres_filas_TF <- rownames(omicas$TF)
nuevos_nombres_filas <- paste0(nombres_filas_TF, 'TF')
rownames(omicas$TF) <- nuevos_nombres_filas

# Define data matrices for each block
omicas_cnv <- omicas$CNV
omicas_methyl <- omicas$Methyl
omicas_miRNA <- omicas$miRNA
omicas_tf <- omicas$TF

# Remove useless variables
rm(assoc, clinical, GeneExpr, nuevos_nombres_filas, nombres_filas_TF)

# Create a list of blocks
blocks <- list("A" = 1:ncol(omicas_cnv), 
               "B" = (ncol(omicas_cnv) + 1):(ncol(omicas_cnv) + ncol(omicas_methyl)),
               "C" = (ncol(omicas_cnv) + ncol(omicas_methyl) + 1):(ncol(omicas_cnv) + ncol(omicas_methyl) + ncol(omicas_miRNA)),
               "D" = (ncol(omicas_cnv) + ncol(omicas_methyl) + ncol(omicas_miRNA) + 1):ncol(omicas_tf))

# Combine all omics into a single matrix
omicas_all <- rbind(omicas_cnv, omicas_methyl, omicas_miRNA, omicas_tf)

# Define X and Y for data splitting
X <- t(omicas_all)
subtypes$subtype <- factor(subtypes$subtype)
Y <- subtypes$subtype

# Split the dataset into training and testing sets
set.seed(123)  # For reproducibility
indices <- createDataPartition(Y, p = 0.8, list = FALSE)
x_train <- X[indices, ]
x_test <- X[-indices, ]
y_train <- Y[indices]
y_test <- Y[-indices]
```

Hyperparameters
-> mtry -> a vector of length equal to the number of blocks, where the m-th entry of the vector
gives the number of variables to sample from block m. The default values are sqrt(p_1) +
sqrt(p_2) + ... sqrt(p_M) where p_m denotes the number of variables in the m-th block (m = 1, ..., M) and sqrt() denoted the square root function -> let the function calculate it

-> splitrule, block.method, replace -> default

-> try values for two hyperparameters: nsets and num.trees:
-> num.trees = c(100, 200)
-> nsets = c(100, 200)

```{r}
# Define the evaluation function for cross-validation
cv_func <- function(num.trees, nsets, folds, x_train, y_train, blocks) {
  cv_results <- lapply(folds, function(fold_index) {
    x_train_fold <- x_train[fold_index[[1]], ]
    y_train_fold <- y_train
    
    # Fit the blockForest model with the current fold
    result <- blockfor(X = t(x_train_fold), y = y_train_fold, blocks = blocks, 
                       num.trees = num.trees, replace = TRUE,
                       nsets = nsets, num.trees.pre = 50, splitrule = "gini", 
                       block.method = "BlockForest", importance='impurity')
  
    return(result)
  })
  
  return(cv_results)
}

# Define the ranges of values for the hyperparameters
param_grid <- expand.grid(
  num.trees = c(100, 200),
  nsets = c(10, 20)
)

# Apply cv_func for each combination of hyperparameters in param_grid
results <- mapply(function(num_trees, nsets) {
  cv_func(num.trees = num_trees, 
          nsets = nsets, 
          folds = folds, 
          x_train = x_train, 
          y_train = y_train, 
          blocks = blocks)
}, param_grid$num.trees, param_grid$nsets)
```

```{r}
# Define a function to get the lowest prediction error from the results
get_min_prediction_error <- function(results) {
  min_errors <- sapply(results, function(result) {
    min(result$forest$prediction.error)
  })
  return(min(min_errors))
}

# Filter the results to remove those with non-finite prediction errors
filtered_results <- results[sapply(results, function(result) all(is.finite(result$forest$prediction.error)))]

# Get the lowest prediction error from all filtered results
min_error <- get_min_prediction_error(filtered_results)

# Find the combination of hyperparameters with the lowest prediction error
best_combination_index <- which.min(sapply(filtered_results, function(result) min(results$forest$prediction.error)))
best_combination <- param_grid[best_combination_index, ]

print(paste("The combination of hyperparameters with the lowest prediction error is:", 
            paste(names(best_combination), best_combination, sep = "=", collapse = ", "), 
            "with a prediction error of", min_error))
```

## Apply model with the best hyperparameters found

```{r}
# Apply the blockfor model with the best parameter combination on the test set
best_model <- blockfor(X = x_train, y = y_train, blocks = blocks, 
                       num.trees = 100, replace = TRUE,
                       nsets = 10, num.trees.pre = 50, splitrule = "gini", 
                       block.method = "BlockForest", importance='impurity')

# Convert the x_test matrix to a dataframe
x_test_df <- data.frame(x_test)

predictions <- predict(object = best_model$forest,
                       data = x_test_df, 
                       predict.all = FALSE, 
                       num.trees = 100, 
                       type = 'response', 
                       num.threads = 1, 
                       verbose = TRUE)
```

```{r}
# Genera la matriz de confusión
conf_matrix <- confusionMatrix(predictions$predictions, y_test)

# Imprime la matriz de confusión
print(conf_matrix)
```

## 10 variables mas importantes
```{r}
# Get the importance of the variables
importancia_variables <- best_model$forest$variable.importance

# Sort the variables by importance from highest to lowest
importancia_ordenada <- importancia_variables[order(importancia_variables, decreasing = TRUE)]

# Filter variables with positive importance
importancia_positiva <- importancia_ordenada[importancia_ordenada > 0]

# Filter variables with negative importance
importancia_negativa <- importancia_ordenada[importancia_ordenada < 0]

# Select the top 10 most important variables
top_10_importancia <- importancia_ordenada[1:11]

## Check the column names in x_train
colnames_x_train <- colnames(x_train)

# Get the names of the top 10 most important variables
top_10_nombres <- names(top_10_importancia)

# Check if the variable names are in x_train
top_10_nombres <- top_10_nombres[top_10_nombres %in% colnames_x_train]

top_10_nombres

# Create a bar plot of the top 10 most important variables with names
barplot(top_10_importancia[top_10_nombres], main = "Top 10 Important Variables",
        xlab = "Importance", ylab = "Variables", names.arg = top_10_nombres, las = 2)
```


## Top 10 least important variables 

```{r}
# Sort the variables by importance from lowest to highest
importancia_ordenada <- importancia_variables[order(importancia_variables)]

# Select the bottom 10 least important variables
bottom_10_importancia <- importancia_ordenada[1:11]

# Get the names of the bottom 10 least important variables
bottom_10_nombres <- names(bottom_10_importancia)

# Check if the variable names are in x_train
bottom_10_nombres <- bottom_10_nombres[bottom_10_nombres %in% colnames_x_train]

# Create a bar plot of the bottom 10 least important variables
barplot(bottom_10_importancia, main = "Bottom 10 Important Variables",
        xlab = "Variables", ylab = "Importance")
```


## Find which block each important variable belongs to

```{r}
library(dplyr)
# Get the names of the variables for each block
names_cnv <- colnames(t(omicas_cnv))
names_methyl <- colnames(t(omicas_methyl))
names_miRNA <- colnames(t(omicas_miRNA))
names_tf <- colnames(t(omicas_tf))

# Create a list with the variable names and their corresponding block
block_names <- list(
  CNV = names_cnv,
  Methylation = names_methyl,
  miRNA = names_miRNA,
  TF = names_tf
)

# Function to find the block to which a variable belongs
find_block <- function(variable, block_names) {
  for (block in names(block_names)) {
    if (variable %in% block_names[[block]]) {
      return(block)
    }
  }
  return(NA)  # Returns NA if not found in any block
}

# Important variables
important_vars <- c("MIR34A", "PRAMEF3", "RERE", "SPATA21", "HES3", "HNRNPCL1", "CLCNKA", "C1orf195", "SNORA59A", "FAM131C", "ALDH4A1", "PLEKHM2" ,  "PAX7" ,     "B3GALT6" ,  "TNFRSF25" , "CASZ1",     "LINC00115", "ARHGEF10L")

# Create a dataframe to store the important variables and their corresponding blocks
important_vars_blocks <- data.frame(
  Variable = important_vars,
  Block = sapply(important_vars, find_block, block_names = block_names)
)

print(important_vars_blocks)
```

## ANOVA

```{r}
# Create a table of the most important variables and their values in the training dataset
top_variables_data <- as.data.frame(x_train[, top_10_nombres])
top_variables_data$subtype <- y_train

# Analyze the association between variables and cancer subtypes using ANOVA
anova_results <- list()

for (variable in top_10_nombres) {
  formula <- as.formula(paste(variable, "~ subtype"))
  anova_results[[variable]] <- summary(aov(formula, data = top_variables_data))
}

# Display ANOVA results
anova_summary <- sapply(anova_results, function(x) x[[1]][["Pr(>F)"]][1])
anova_summary <- data.frame(Variable = names(anova_summary), PValue = anova_summary)
anova_summary <- anova_summary %>% arrange(PValue)

print("ANOVA results for the most important variables:")
print(anova_summary)
```

